---
title: "Time Series (ARIMA) Prediction"
author: '"Popoola Olawale"'
date: "2023-07-11"
output: html_document
---
 
      __PREDICTING THE NUMBER OF PICKUP RIDES FOR YELLOW TAXI IN NEW YORK CITY_
                              __MARCH 2023__
        _HOURLY PREDICTION FOR NEWYORK YELLOW TAXI FOR THE MONTH OF MARCH 2023_
                              
                              
The issue of demand affects most businesses and it is the same for the transportation business. In this work, I attempt to build a model that uses the daily number of pickup rides for yellow taxi in New York city to estimate demand. For this work, I'll attempt to use ARIMA model to predict yellow taxi demands by riders. The data set contains the hourly picked up rides for the month of March by New York Yellow taxis.


__Data Preprocessing__

The original file wad downloaded from https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page. The file contains 3403766 rows and 19 columns.


```{r}
library(arrow)
library(tidyverse)
library(forecast)
library(tseries)
```


 Feature Engineering to deduce the amount of car pickup per hour from the original file and a new file was created. this file will be used for the analysis. 
 
This is the code.
 a <- nyc_ytaxi_march_2023 %>% 
  distinct(tpep_pickup_datetime) %>% 
  transmute(Hourly = floor_date(tpep_pickup_datetime, unit = "hour")) %>% 
  count(Hourly)
  
saveRDS(a, file = "nyc_ytaxi_march_2023.RDS")
write.csv(nyc_ytaxi_march_2023, file= "nyc_ytaxi_march_2023.csv")
 
 
 I load the newly created file
```{r}
ytaxi_march_2023 <- read.csv("nyc_ytaxi_march_2023.csv")
```
 
 
_Data Cleaning_

The first and last rows contain inaccurate data from previous and following months. I will delete these rows as they are not necessary for this analysis.
```{r}
ytaxi_march_2023 <- ytaxi_march_2023[-c(1:26,770:801),]

dim(ytaxi_march_2023)
class(ytaxi_march_2023$n)
```
 
 
_Data Conversion_
```{r}
a <- ts(ytaxi_march_2023$n, start = 1, frequency = 24)
```

_Exploratory Analysis_
```{r}
summary(ytaxi_march_2023$n)

plot(a)
abline(reg = lm(a~time(a)))

boxplot(a~cycle(a))
```

From the descriptive analysis, we can see there is no trend in the data set as the mean appears to be constant. However, there is some variance in the data set. From the box plot, we can see that the first hours of the data has different variation and the mean average also varies. While from 11am to 8pm, we see less variance in the data set.


__Data Analysis__

# Finding model parameters for ARIMA

```{r}
plot(acf(a))  
plot(pacf(a))
```

AR I MA
 p =1, d = 0, q =6
 
The moving average value (q) invalidated the model, so I attempted the q value of 1 to derive a model. 

Lets fit the Arima model and predict for the next 3 days

```{r}
fit <- arima(a, order = c(1,0,1), seasonal = list(order = c(1,0,1), period = 24))

pred <- predict(fit,n.ahead = 3*24)
names(pred)
results <- pred$pred
round(results)

```

The results are fair when we account for the fact that the first value in the dataset is most likely not true.

```{r}
print(summary(fit))
checkresiduals(fit)
```

From the residuals, we can see that the distribution of the data in the histogram is okay in relation to Gaussian distribution and the confidence interval is okay. We can the say the model is a good fit for the data.



Standard deviation is square root of the sigma value. The standard deviation shows that variance of the predicted results.
```{r}
sd <- sqrt(28184)
sd
```

# I attempt a second model
```{r}
plot(acf(diff(a)))
plot(pacf(diff(a)))
```

When differentiation was added to this model, there was no output for the model but I change the value of the partial acf.

```{r}
fit1 <- arima(a, order = c(1,1,4), seasonal = list(order = c(1,1,4), period = 24))


pred1 <- predict(fit1,n.ahead = 3*24)
names(pred1)
results1 <- pred1$pred
round(results1)
```
Similar as the first model, taking into account the degree of reliability of the first value of the month, the model results seems okay.



```{r}
print(summary(fit1))
checkresiduals(fit1)
```

From the residuals, we can see that the distribution of the data is okay in relation to Gaussian distribution and the confidence interval is okay. We can the say the model is a good fit for the data.


Standard Deviation is sqrt sigma
```{r}
sd1 <- sqrt(20754)
sd1
```


The Standard deviation of the second model is better than the first model. I would like to improve the accuracy of my prediction for the analysis in subsequent model construction.
